The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2013-08-31 floats/figures.tex
Rework discussion and add one figure
General reread and polish

% fig:locmap
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=8cm]{cordillera-climate-locmap}
	\end{center}
	\caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) \citep{dyke-2004}. The background map consists of ETOPO1 (\cite{data:etopo1}) and Natural Earth Data (\cite{data:naturalearth}).}
	\label{fig:locmap}
\end{figure}

% fig:topo
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-topo}
	\end{center}
	\caption{Topography maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis data used as a reference for temperature lapse-rate corrections, and topography map from ETOPO1 used as basal topography.}
	\label{fig:topo}
\end{figure*}

% fig:temp
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-temp}
	\end{center}
	\caption{Summer (JJA) and winter (DJF) temperature maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis climatologies.}
	\label{fig:temp}
\end{figure*}

% fig:prec
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-prec}
	\end{center}
	\caption{Summer (JJA) and winter (DJF) precipitation maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis climatologies.}
	\label{fig:prec}
\end{figure*}

% fig:cool05
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-cool05}
	\end{center}
	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 5\degC colder than present for each climate forcing.}
	\label{fig:cool05}
\end{figure*}

% fig:extent
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-extent}
	\end{center}
	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
	\label{fig:extent}
\end{figure*}

% fig:ivolarea
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
	\end{center}
	\julien[inline]{I will make the figure readable in black-and-white and highlight markers corresponding to the ``best'' runs.}
	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
	\label{fig:ivolarea}
\end{figure}

% fig:best
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-best}
	\end{center}
	\caption{Ice surface topography (black contours every 1000\,m) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing. 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) ice margin (blue line) from \citet{dyke-2004}.}
	\label{fig:best}
\end{figure*}

>>>>

% fig:locmap
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=8cm]{cordillera-climate-locmap}
	\end{center}
    \irina[inline]{include all geographic names}
    \arjen[inline]{elevation scale white-end confused with glaciers}
    \arjen[inline]{remove Arctic circle and miles scale}
    \arjen[inline]{use CIS, LIS and IIS for ice sheet names}
    \arjen[inline]{use N, S, E and W instead of minus, add 75\degree N label}
	\caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) \citep{dyke-2004}. The background map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth}. Assembled with GRASS~GIS \citep{soft:grass} and Inkscape \citep{web:inkscape}}
	\label{fig:locmap}
\end{figure}

% fig:topo
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-topo}
	\end{center}
	\caption{Topography maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis data used as a reference for temperature lapse-rate corrections, and topography map from ETOPO1 used as basal topography. Drawn using Matplotlib \citep{soft:mpl}.}
	\label{fig:topo}
\end{figure*}

% fig:temp
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-temp}
	\end{center}
	\caption{Summer (JJA) and winter (DJF) temperature maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis climatologies. Drawn using Matplotlib \citep{soft:mpl}.}
	\label{fig:temp}
\end{figure*}

% fig:prec
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-prec}
	\end{center}
	\caption{Summer (JJA) and winter (DJF) precipitation maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis climatologies. Drawn using Matplotlib \citep{soft:mpl}.}
	\label{fig:prec}
\end{figure*}

% fig:cool05
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-cool05}
	\end{center}
	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 5\degC colder than present for each climate forcing. Drawn using Matplotlib \citep{soft:mpl}.}
	\label{fig:cool05}
\end{figure*}

% fig:extent
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-extent}
	\end{center}
	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing. Drawn using Matplotlib \citep{soft:mpl}.}
	\label{fig:extent}
\end{figure*}

% fig:ivolarea
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
	\end{center}
	\julien[inline]{I will make the figure readable in black-and-white and highlight markers corresponding to the ``best'' runs.}
	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing. Drawn using Matplotlib \citep{soft:mpl}.}
	\label{fig:ivolarea}
\end{figure}

% fig:best
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-best}
	\end{center}
	\caption{Ice surface topography (black contours every 1000\,m) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing. 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) ice margin (blue line) from \citet{dyke-2004}. Drawn using Matplotlib \citep{soft:mpl}.}
	\label{fig:best}
\end{figure*}

% fig:oroprecip
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=8cm]{cordillera-climate-oroprecip}
	\end{center}
	\caption{Schematic representation of the orographic precipitation effect over the Coast Mountains. In a GCM of lower resolution, the precipitation peak appear smoother and the precipitation shadow is not as pronounced. Drawn using Inkscape \citep{web:inkscape}.}
	\label{fig:oroprecip}
\end{figure}

### 2013-08-31 sections/climate.tex
General reread and polish

WorldClim is a high-resolution climate dataset built from meteorological observations \citep{data:worldclim}. It was built by spatial interpolation over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}, between a large set of measurements taken at weather stations worldwide. This provides a resolution much higher than attained through GCM in areas where sufficient amount of observations exist.

@@

In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global climate reanalyses and one regional climate reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis \citep[CFSR,][]{data:cfsr}, and the North American Regional Reanalysis \citep[NARR,][]{data:narr}. Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).

@@

WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}. In addition, WorldClim data was extrapolated in a nearest-neighbour fashion to cover oceans grid points using the interpolate.griddata method from Scipy \citep{soft:scipy}. Note that we choose to present here (Fig.~\ref{fig:temp} and~\ref{fig:prec}) the original rather than interpolated data in order to highlight differences related to spatial resolution among datasets.

>>>>

WorldClim is a high-resolution climate dataset built from meteorological observations \citep{data:worldclim}. It was built by spatial interpolation over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}, between a large set of measurements taken at weather stations worldwide. This provides a resolution much higher than attained by GCMs in areas where sufficient amount of observations exists.

@@

In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global climate reanalyses and one regional climate reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and the North American Regional Reanalysis (NARR) \citep{data:narr}. Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).

@@

WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}. In addition, WorldClim data was extrapolated in a nearest-neighbour fashion to cover oceans grid points using the interpolate.griddata method from Scipy \citep{soft:scipy}. Note that we choose to present here the original rather than interpolated data in order to highlight differences related to spatial resolution among datasets (Fig.~\ref{fig:temp} and~\ref{fig:prec}).

### 2013-08-31 sections/discuss.tex
General reread and polish

\julien[noline]{This sentence may be better placed in the results section.}
We compare the numerically modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating. This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than in the Laurentide ice sheet \citep{dyke-2004}.
Although the outcome of our numerical simulations strongly depends on which dataset is used as a climate forcing, common discrepancies between the modelled ice sheet geometry and the ice margin mapped by \citet{dyke-2004} can be observed (Fig.~\ref{fig:cool05}-\ref{fig:best}).

For all climate forcing used, the modelled eastern margin of the ice sheet extends further east than the reconstructed boundary between the Cordilleran and the Laurentide ice sheets (Fig.~\ref{fig:best}). This can be explained by the potential effects of the growing ice-sheet on regional climate not included in our model, or by buttressing effects against the Laurentide Ice Sheet. The Cordilleran ice sheet initiated from the junction of mountain ice caps over the major reliefs~\citep{clague-1989}. In our simulations, a continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain. This continuous ice cover may have enhanced the topographical barrier already formed by the western ranges, resulting in fewer precipitation and warmer air in the interior. \citet{langen-etal-2012} demonstrated that this process, which they refer to as a ``self-inhibiting growth'', may have been limiting on the build-up of the Greenland Ice Sheet. Alternatively, the Laurentide Ice Sheet, not included in our model, may have formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.

Our simulations produce anomalous ice cover on parts of the continental shelf in the Arctic Ocean. These regions experience a marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}). However in our simulations sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land. A similar effect occurs over Great Bear Lake in some simulations, particularly those driven by the NCEP/NCAR forcing (Fig.~\ref{fig:extent}). This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.

@@

Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data shows that ice cover was sparse. Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions (Fig.~\ref{fig:topo}, Table~\ref{tab:reanalyses}).

As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid. However the ability of a GCM to reproduce these contrasts in precipitation is bound to its resolution. In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}). This interpretation does not apply to the CFSR, which exhibits high inland precipitation rates despite of its high resolution (Table~\ref{tab:reanalyses}).

@@

Some simplifications were made in the surface mass-balance model. Our PDD model do not include refreezing. Refreezing of melted snow and ice and its retention in the snow pack, however, can large alter mass-balance at the surface of an ice sheet \citep{janssens-huybrechts-2000}. Furthermore, we model temperature variability using a constant, homogeneous value of temperature standard deviation. This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inreview}, particularly over regions with such various climates as the Northern American Cordillera.

>>>>

We compare the numerically modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating. This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than in the Laurentide ice sheet \citep{dyke-2004}.
Although the outcome of our numerical simulations strongly depends on which dataset is used as a climate forcing, common discrepancies between the modelled ice sheet geometry and the ice margin mapped by \citet{dyke-2004} can be observed (Fig.~\ref{fig:cool05}-\ref{fig:best}).

For all climate forcing used, the modelled eastern margin of the ice sheet extends further east than the reconstructed boundary between the Cordilleran and the Laurentide ice sheets (Fig.~\ref{fig:best}). This can be explained by the potential effects of the growing ice-sheet on regional climate not included in our model, or by buttressing effects against the Laurentide Ice Sheet. The Cordilleran ice sheet initiated from the junction of mountain ice caps over the major reliefs~\citep{clague-1989}. In our simulations, a continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain. This continuous ice cover may have enhanced the topographical barrier already formed by the western ranges, resulting in fewer precipitation and warmer air in the interior. \citet{langen-etal-2012} demonstrated that this process, which they refer to as a ``self-inhibiting growth'', may have been limiting during the build-up of the Greenland Ice Sheet. Alternatively, the Laurentide Ice Sheet, not included in our model, may have formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.

Our simulations produce anomalous ice cover on parts of the continental shelf in the Arctic Ocean. These regions experience a marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}). However in our simulations, sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land. A similar effect occurs over Great Bear Lake in some simulations, particularly those driven by the NCEP/NCAR forcing (Fig.~\ref{fig:extent}). This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.

@@

Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data shows that the ice cover was sparse. Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions (Fig.~\ref{fig:topo}, Table~\ref{tab:reanalyses}).

As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid. However the ability of a GCM to reproduce these contrasts in precipitation is bound to its resolution. In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}). This interpretation does not apply to the CFSR climatology, which exhibits high inland precipitation rates despite of the high model resolution (Table~\ref{tab:reanalyses}).

@@

Furthermore, some simplifications were made in the surface mass balance model. Our PDD model do not include refreezing. Refreezing of melted snow and ice and retention in the snow pack, however, can largely alter mass balance at the surface of an ice sheet \citep{janssens-huybrechts-2000}. Additionally, we model temperature variability using a constant, homogeneous value of temperature standard deviation. This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inreview}, particularly over regions with such various climates as the Northern American Cordillera.

### 2013-08-31 sections/intro.tex
General reread and polish

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to model ice sheets in Northern America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}. While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism}\julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM. We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}. To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since the one by \citet{robert-1991}. While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, ours encompasses most of the LGM ice cover and adjacent non-glaciated parts of northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}). Our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response following the large advances made in scientific computing and ice sheet modelling since then. Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}. Results are exposed in section~\ref{sec:results} and discussed in section~\ref{sec:discussion}.

>>>>

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to simulate ice sheets in Northern America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}. While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism}\julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM. We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}. To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since the one by \citet{robert-1991}. While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, ours encompasses most of the LGM ice extent and adjacent non-glaciated parts of northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}). Our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response following the large advances made in scientific computing and ice sheet modelling since then. Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}. Results are exposed in section~\ref{sec:results} and discussed in section~\ref{sec:discussion}.

### 2013-08-31 sections/model.tex
General reread and polish

We use PISM, a Parallel Ice Sheet Model, version stable~0.5.12 \citep[e.g.,][]{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012,web:pism}. Given basal topography and climate forcing, the model computes the thermal and dynamic state of an ice sheet and the associated lithospheric response. Our modelling domain encompasses most of the area covered by the Cordilleran Ice Sheet at LGM except of the western Alaskan ranges and includes non-glaciated regions in northern Yukon Territory and interior Alaska. Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.

@@

SSA velocities are determined by a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescribed till friction angle $\phi$. We apply a range of till friction angles between 10 and 30\,\degree, where the lower friction angles occur at low elevation to mimic the more deformable beds associated with presence of marine sediments:
\julien{phi here is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}

@@

where the integration interval $[t_0; t_1]$ is one week and $\mathrm{erfc}$ is the complementary error function. The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\degC, which is the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} similarly to \citet{seguinot-inreview}. The ablation model incorporates degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Prior to surface mass balance computation, the model dynamically applies a lapse-rate correction to surface air temperature. This correction accounts for evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the over hand. It uses a reference topography distinct from the basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}). An annual lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations. No lapse-rate corrections applies to precipitation.

As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM, we mimic LGM conditions by applying constant temperature offsets homogeneously over the modelling domain. Each simulation starts from ice-free conditions and runs for 10\,kyr, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.

>>>>

We use PISM, a Parallel Ice Sheet Model, version stable~0.5.12 (e.g.,) \citep{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012,web:pism}. Given basal topography and climate forcing, the model computes the thermal and dynamic state of an ice sheet and the associated lithospheric response. Our modelling domain encompasses most of the area covered by the Cordilleran Ice Sheet at LGM except of the western Alaskan ranges and includes non-glaciated regions in northern Yukon Territory and interior Alaska. Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.

@@

SSA velocities are determined by a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescribed till friction angle $\phi$. We apply a range of till friction angles between 10 and 30\,\degree, where the lower friction angles occur at low elevation to mimic the more deformable beds associated with presence of marine sediments:
\needref
\julien{phi is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}

@@

where the integration interval $[t_0; t_1]$ is one week and $\mathrm{erfc}$ is the complementary error function. The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\degC, which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inreview}. The ablation model incorporates degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Prior to surface mass balance computation, the model dynamically applies a lapse-rate correction to surface air temperature. This correction accounts for evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the over hand. It uses a reference topography distinct from the basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}). An annual lapse-rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations. No lapse-rate corrections applies to precipitation.

As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain. Each simulation starts from ice-free conditions and runs for 10\,kyr, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.

### 2013-08-31 sections/results.tex
General reread and polish

Across the range of temperature offsets used, final glaciated area and final ice volume differs widely between different input climatologies (Fig.~\ref{fig:ivolarea}). Using a 5\degC temperature offset (Fig.~\ref{fig:cool05}) final glaciated area differs between WorldClim and CFSR forcing by a factor of 6 and final volume by a factor of 12 (Fig.~\ref{fig:ivolarea}). For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volume. Yet regional patterns are different (Fig.~\ref{fig:extent}).

Under present climate (0\degC offset), WorldClim, NARR and ERA-Interim forcing produce local ice caps in areas presently glaciated. This results are fairly consistent with contemporary ice coverage, although our 10~km resolution, and the occurrence of glaciers and ice caps in the ETOPO1 input basal topography do not allow for a detailed comparison. CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice-sheet growth, which is largely inconsistent with the present state of glaciation. At the opposite side of the temperature offset range, it can be noted that NCAR and CFSR simulations produce oversized ice sheets whose extent is primarily restricted by the domain boundaries.

>>>>

Across the range of temperature offsets used, final glaciated area and final ice volume differs widely between different input climatologies (Fig.~\ref{fig:ivolarea}). Using a 5\degC temperature offset (Fig.~\ref{fig:cool05}) final glaciated area differs between WorldClim and CFSR forcing by a factor of 6 and final volume by a factor of 12 (Fig.~\ref{fig:ivolarea}). For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volume, yet regional patterns are different (Fig.~\ref{fig:extent}).

Under present climate (0\degC offset), WorldClim, NARR and ERA-Interim forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}). This results are fairly consistent with contemporary ice coverage, although both the 10~km horizontal resolution, and the occurrence of glaciers and ice caps in the ETOPO1 input basal topography prevent a detailed comparison. CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice-sheet growth, which is largely inconsistent with the present state of glaciation. At the opposite side of the temperature offset range, it can be noted that NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extent is primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}).
